A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-10-12 doc/paper1-submission/paper4.tex
Incorporate Emily's comments.

These optimality criteria help to define what it means for the supertree
to represent the input trees, as well as justifying and explaining
why various features of the supertree exist. The pipeline described
below produces a supertree that satisfies the first three optimality
criteria and is a greedy approximation of a solution to the fourth
goal. It is not guaranteed to display as many input nodes as possible.
Even if the summary tree does accomplish goal 4, it is not necessarily
a \emph{unique} optimum. The pipeline takes a greedy approach to producing
a summary tree by attempting to add groupings from the trees in order
of the trees ranking. This can be viewed as a greedy solution to the
problem of finding the tree with the maximum sum of displayed groups'
weighted scores criterion (described in the appendix A) where the
weights from the trees are so extreme that displaying one group from
a highly ranked tree is preferred to displaying all of the groupings
from lower ranked trees.

@@

After the preprocessing steps, the inputs have been converted to a
set of rooted phylogenetic estimates in which each leaf is mapped
to a terminal taxon in the exemplified taxonomic tree. The goal of
the remainder of the pipeline is to construct a tree that maximizes
the sum displayed groups' weighted scores criterion. This is accomplished
in four steps: (1) dividing the full problem into subproblems based
on uncontested taxa, (2) constructing a summary solution for each
subproblem by greedily creating a maximally-sized list of groupings
that can all be displayed simultaneously; (3) grafting the subproblem
solutions into a single supertree; and (4) grafting (or ``unpruning'')
the taxonomy-only taxa onto the solution to produce a complete summary
tree.

@@

For the sake of efficiency, propinquity uses a divide-and-conquer
approach to construct the supertree. Subproblems are identified by
searching through the taxonomy tree to find any taxa are not contested
by any single input tree. Here we say that input tree $T_{i}$ contests
taxon $x$ in the pruned taxonomy, if $x$ is not monophyletic in
any resolution of tree $T_{i}$. Thus, polytomies in an input tree
are treated as soft polytomies, and a taxon is not contested merely
because it is not displayed by an input tree.

@@

The BUILD algorithm as originally stated by \citet{AhoSSU1981} applies
to a collection of rooted triplets. Instead of decomposing each input
split into a collection of rooted triplets, we instead modify the
BUILD algorithm to apply directly to larger rooted splits. The modified
BUILD algorithm constructs a tree compatible with a collection of
rooted splits, and returns failure if such a tree does not exist.
This modified algorithm recovers the original BUILD algorithm if only
rooted triplets are supplied as input. When larger splits are supplied
as input, the results are the same as if each was was decomposed into
all implied triplets. The modified build algorithm has order $O(N^{2}+N^{2}E+NL)$
where $N$ is the number of splits passed in, $E$ is the average
size of the exclude group, and $L$ is the total number of leaves.
This simplifies to $O(N^{2})$ if all splits are triplets. In this
approach splits are either entirely retained or entirely discarded
- consistent rooted triplets from conflicting splits are not retained.
However, when unpruning taxonomy-only taxa (see below), we make an
attempt to break ties in a way that preserves some partial information
from conflicting splits by attaching taxa from conflicting splits
at their common ancestor. Figure \ref{fig:subproblem-solutions.}
shows the solutions that would be obtained by applying our modified
version of the BUILD algorithm to the subproblems show in figure \ref{fig:subproblems-example}.

@@

However, a taxon may be incompatible with the grafted solution; we
refer to such taxa as ``broken taxa.'' If a broken taxon contains
some unsampled children, it is not clear where these unsampled children
should be attached to the grafted solution. One approach would be
to mimic the application of apply Algorithm \ref{alg:ConsistentSplitsFromRankedList}
to the full (unpruned) taxonomic tree. This would be equivalent to
collapsing each edge in the taxonomy that attaches a broken taxon
to its parent. The unsampled children of broken taxa would attach
at their least inclusive ancestral taxon which is unbroken. In cases
of several adjacent taxa are broken, this can lead to polytomies of
very high degree deep in the tree. This can make the summary tree
difficult to navigate. Thus, we have adopted an alternative solution.
The \texttt{otc-unprune-solution-and-name-unnamed-nodes} tool from
otcetera attaches the unsampled children of a broken taxa to the grafted
solution as children of the MRCA of the sampled children.

@@

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.38]{example3__cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.38]{example3__grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.38]{example3__unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.38]{example3__unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R and instead attached at the MRCA of R1, R2, and
R3.}
\end{figure}

@@

In order to annotate each node in the summary supertree, it is first
necessary that each node have a unique identifier. Nodes whose include
group correspond exactly to the include group of a node in the taxonomy
are given the same identifier as the corresponding taxonomy node.
These identifier are of the form $ottX$ where $X$ is an integer
OTT id. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
ids for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
is the numerically smallest OTT ID that is a descendant of $n$, and
$X_{2}$ is the next the smallest ID that can be chosen to designate
$n$ as the MRCA. Because new taxa added to OTT will be given higher
OTT Ids, the use of the lowest numbered OTT Ids as designators increases
the chance that a node label can be encountered in a subsequent version
of the tree (though the taxonomic content may change). The deterministic
choice of designators also makes the labeling insensitive to branch
rotation of the grafted solution tree.

@@

The tool creates a multigraph by starting with a graph is isomorphic
to the taxonomic tree. The nodes and edges created in this step will
be referred to as the ``taxonomic graph.'' Next, we add nodes and
edges to that graph in a procedure that we refer to as ``embedding''
the input trees into the taxonomy. A node is introduced for each node
in an input tree, and these nodes are mapped the MRCA nodes in the
taxonomic graph. In other words, for any node $y$ in an input tree
with a cluster of descendants, $\mathcal{C},$ we find the most tipward
node $z$ in the taxonomic graph that is an ancestor to all of the
taxa in $\mathcal{C}$; let $m(y)=z$ refer to this mapping, and $m^{\prime}(z)=y$
refer to the reverse mapping. Each edge $e_{ij}$ in a source tree
$i$ connects ancestor to its descendant, $a(e_{ij})\rightarrow d(e_{ij})$.
The edges are introduced into the graph. We also introduce new edges
to create a from $m(a(e_{ij}))$ through its descendants to $m(d(e_{ij}))$;
we denote this path $p(e_{ij})$ and refer to the edges in the path
as ``embedding edges for tree $i$''. Note, that this is a path
through the taxonomic nodes, while the edge $e_{ij}$ connects source
tree nodes. The mapping between $e_{ij}$ and $p(e_{ij})$ is stored,
and the edges are labelled with the index $i$ so that it is clear
which tree created them. Because the taxonomic tree is highly unresolved,
it is frequently the case that $m(a(e_{ij}))$ is the same node as
$m(d(e_{ij}))$; in these cases the embedding edge is a loop. This
situation occurs whenever edge $e_{ij}$ can resolve part of the polytomy
represented by a taxon.

@@

After every input tree and the taxonomy tree have been embedded into
the taxonomic graph, we perform postorder traversal over the taxonomic
graph that underlies the multi-graph. For any internal node (each
of which corresponds to a non-terminal taxon) we determine whether
or not it is contested by examining each input tree. We can determine
tree $i$ contests the taxon represented by taxonomic node $x$ by
looking at the parents of all of the ``exiting'' embedding edges
for tree $i$. These are the set of embedding edges that have $x$
as a daughter and have a parent node that is not $x$ (ergo a parent
node that is taxonomically higher than $x$). If there are more than
one parent nodes in this set of exiting embedding edges, then tree
$i$ contests that taxon. If there is only parent node, then the all
of the constituent taxa belonging to this taxon that are present in
tree $i$ have one parent that is more inclusive; this means that
the input tree does not contest monophyly of the taxon.

>>>>

These optimality criteria help to define what it means for the supertree
to represent the input trees, as well as justifying and explaining
why various features of the supertree exist. The pipeline described
below produces a supertree that satisfies the first three optimality
criteria and is a greedy approximation of a solution to the fourth
goal. It is not guaranteed to display as many input nodes as possible.
Even if the summary tree does accomplish goal 4, it is not necessarily
a \emph{unique} optimum. The pipeline takes a greedy approach to producing
a summary tree by attempting to add groupings from the trees in order
of the trees ranking. This can be viewed as a greedy solution to the
problem of finding the tree with the maximum sum of displayed groups'
weighted scores criterion (MSDGWS, described in the appendix A) where the
weights from the trees are so extreme that displaying one group from
a highly ranked tree is preferred to displaying all of the groupings
from lower ranked trees.

@@

After the preprocessing steps, the inputs have been converted to a
set of rooted phylogenetic estimates in which each leaf is mapped
to a terminal taxon in the exemplified taxonomic tree. The goal of
the remainder of the pipeline is to construct a tree that maximizes
the sum of displayed groups' weighted scores criterion. This is accomplished
in four steps: (1) dividing the full problem into subproblems based
on uncontested taxa, (2) constructing a summary solution for each
subproblem by greedily creating a maximally-sized list of groupings
that can all be displayed simultaneously; (3) grafting the subproblem
solutions into a single supertree; and (4) grafting (or ``unpruning'')
the taxonomy-only taxa onto the solution to produce a complete summary
tree.

@@

For the sake of efficiency, propinquity uses a divide-and-conquer
approach to construct the supertree. Subproblems are identified by
searching through the taxonomy tree to find any taxa that are not contested
by any single input tree. Here we say that input tree $T_{i}$ contests
taxon $x$ in the pruned taxonomy, if $x$ is not monophyletic in
any resolution of tree $T_{i}$. Thus, polytomies in an input tree
are treated as soft polytomies, and a taxon is not contested merely
because it is not displayed by an input tree.

@@

The BUILD algorithm as originally stated by \citet{AhoSSU1981} applies
to a collection of rooted triplets. Instead of decomposing each input
split into a collection of rooted triplets, we instead modify the
BUILD algorithm to apply directly to larger rooted splits. The modified
BUILD algorithm constructs a tree compatible with a collection of
rooted splits, and returns failure if such a tree does not exist.
This modified algorithm recovers the original BUILD algorithm if only
rooted triplets are supplied as input. When larger splits are supplied
as input, the results are the same as if each was was decomposed into
all implied triplets. The modified build algorithm has order $O(N^{2}+N^{2}E+NL)$
where $N$ is the number of splits passed in, $E$ is the average
size of the exclude group, and $L$ is the total number of leaves.
This simplifies to $O(N^{2})$ if all splits are triplets. In this
approach splits are either entirely retained or entirely discarded
- consistent rooted triplets from conflicting splits are not retained.
However, when unpruning taxonomy-only taxa (see below), we make an
attempt to break ties in a way that preserves some partial information
from conflicting splits by attaching taxa from conflicting splits
at their common ancestor. Figure \ref{fig:subproblem-solutions.}
shows the solutions that would be obtained by applying our modified
version of the BUILD algorithm to the subproblems shown in figure \ref{fig:subproblems-example}.

@@

However, a taxon may be incompatible with the grafted solution; we
refer to such taxa as ``broken taxa.'' If a broken taxon contains
some unsampled children, it is not clear where these unsampled children
should be attached to the grafted solution. One approach would be
to mimic the application of Algorithm \ref{alg:ConsistentSplitsFromRankedList}
to the full (unpruned) taxonomic tree. This would be equivalent to
collapsing each edge in the taxonomy that attaches a broken taxon
to its parent. The unsampled children of broken taxa would attach
at their least inclusive ancestral taxon which is unbroken. In cases
of several adjacent taxa are broken, this can lead to polytomies of
very high degree deep in the tree. This can make the summary tree
difficult to navigate. Thus, we have adopted an alternative solution.
The \texttt{otc-unprune-solution-and-name-unnamed-nodes} tool from
otcetera attaches the unsampled children of a broken taxa to the grafted
solution as children of the MRCA of the sampled children.

@@

\begin{figure}
\begin{centering}
\subfloat[Taxonomy]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.38]{example3__cleaned_ott-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}\subfloat[Grafted solution]{\begin{centering}
\begin{minipage}[c][1\totalheight][t]{0.45\columnwidth}%
\begin{center}
\includegraphics[scale=0.38]{example3__grafted_solution-labels\lyxdot tre}
\par\end{center}%
\end{minipage}
\par\end{centering}
}\hfill{}
\par\end{centering}
\begin{centering}
\subfloat[Unpruned tree - broken taxa removed]{\begin{centering}
\includegraphics[scale=0.38]{example3__unprune-old-labels\lyxdot tre}
\par\end{centering}
}\hfill{}\subfloat[Unpruned tree with MRCA-attachment]{\begin{centering}
\includegraphics[scale=0.38]{example3__unprune-new-labels2\lyxdot tre}
\par\end{centering}
}
\par\end{centering}
\caption{\label{fig:Two-approaches-to-unpruning}Two approaches to unpruning.
Taxa G and R in the taxonomy (a) are broken because they conflict
with the grafted solution (b). Removing these broken taxa from the
taxonomy before unpruning leads to taxa R4, R5, and R6 being attached
directly at taxon N, as in tree (c). In tree (d), the children of
the broken taxon R are instead attached at the MRCA of R1, R2, and
R3. Our method follows the second approach.}
\end{figure}

@@

In order to annotate each node in the summary supertree, it is first
necessary that each node have a unique identifier. Nodes whose include
group correspond exactly to the include group of a node in the taxonomy
are given the same identifier as the corresponding taxonomy node.
These identifiers are of the form $ottX$ where $X$ is an integer
OTT id. We generate a label of the form $mrcaottX_{1}ottX_{2}$ for
an non-taxonomic node $n$ where $X_{1}$ and $X_{2}$ are the OTT
ids for two leaves, $n$ is the MRCA of these leaves, and $X_{1}$
is the numerically smallest OTT ID that is a descendant of $n$, and
$X_{2}$ is the next the smallest ID that can be chosen to designate
$n$ as the MRCA. Because new taxa added to OTT will be given higher
OTT Ids, the use of the lowest numbered OTT Ids as designators increases
the chance that a node label can be encountered in a subsequent version
of the tree (though the taxonomic content may change). The deterministic
choice of designators also makes the labeling insensitive to branch
rotation of the grafted solution tree.

@@

The tool creates a multigraph by starting with a graph isomorphic
to the taxonomic tree. The nodes and edges created in this step will
be referred to as the ``taxonomic graph.'' Next, we add nodes and
edges to that graph in a procedure that we refer to as ``embedding''
the input trees into the taxonomy. A node is introduced for each node
in an input tree, and these nodes are mapped the MRCA nodes in the
taxonomic graph. In other words, for any node $y$ in an input tree
with a cluster of descendants, $\mathcal{C},$ we find the most tipward
node $z$ in the taxonomic graph that is an ancestor to all of the
taxa in $\mathcal{C}$; let $m(y)=z$ refer to this mapping, and $m^{\prime}(z)=y$
refer to the reverse mapping. Each edge $e_{ij}$ in a source tree
$i$ connects ancestor to its descendant, $a(e_{ij})\rightarrow d(e_{ij})$.
The edges are introduced into the graph. We also introduce new edges
to create a from $m(a(e_{ij}))$ through its descendants to $m(d(e_{ij}))$;
we denote this path $p(e_{ij})$ and refer to the edges in the path
as ``embedding edges for tree $i$''. Note, that this is a path
through the taxonomic nodes, while the edge $e_{ij}$ connects source
tree nodes. The mapping between $e_{ij}$ and $p(e_{ij})$ is stored,
and the edges are labelled with the index $i$ so that it is clear
which tree created them. Because the taxonomic tree is highly unresolved,
it is frequently the case that $m(a(e_{ij}))$ is the same node as
$m(d(e_{ij}))$; in these cases the embedding edge is a loop. This
situation occurs whenever edge $e_{ij}$ can resolve part of the polytomy
represented by a taxon.

@@

After every input tree and the taxonomy tree have been embedded into
the taxonomic graph, we perform postorder traversal over the taxonomic
graph that underlies the multi-graph. For any internal node (each
of which corresponds to a non-terminal taxon) we determine whether
or not it is contested by examining each input tree. We can determine
tree $i$ contests the taxon represented by taxonomic node $x$ by
looking at the parents of all of the ``exiting'' embedding edges
for tree $i$. These are the set of embedding edges that have $x$
as a daughter and have a parent node that is not $x$ (ergo a parent
node that is taxonomically higher than $x$). If there are more than
one parent nodes in this set of exiting embedding edges, then tree
$i$ contests that taxon. If there is only one parent node, then the all
of the constituent taxa belonging to this taxon that are present in
tree $i$ have one parent that is more inclusive; this means that
the input tree does not contest monophyly of the taxon.
